### 2014-12-03 writing/F'14 paper/intro.tex
Intro update and misc paper cleanup

To date, experimental investigations of these concepts using
distributed representations have been largely confined to short
phrases \cite{Mitchell:Lapata:2010,Grefenstette-etal:2011,baroni2012entailment}.
However, for robust natural language understanding, it is essential to
model these phenomena in their full generality on complex linguistic
structures. Recent work on the algebraic approach of
\newcite{ClarkCoeckeSadrzadeh2011} has yielded rich frameworks 
for computing the meanings of complex fragments of natural
language compositionally from vector or tensor representations, but
how to learn these representations entirely from data in a typical
machine learning setting remains an open question.

@@

In our first set of experiments, we use this pre-specified logical
grammar to generate controlled data sets encoding the semantic
relationships between pairs of expressions and evaluate whether each
of two classes of neural model---plain recursive NNs and recursive neural
tensor networks (RNTNs; \citealt{socher2013acl1})---can learn those
relationships correctly. In our first experiment
(Section~\ref{sec:join}), we evaluate the ability of these models to
learn the core relational algebra of natural logic from data. Our
second experiment (Section~\ref{sec:recursion}) extends this
evaluation to relations between complex recursive structures like
$(a \plor b)$ and $\plneg (\plneg a \pland \plneg b)$, and our third
experiment (Section~\ref{sec:quantifiers}) involves relations between
quantified statements like \ii{every reptile walks} and \ii{no turtle
  moves}. \tbchecked{We find that the plain recursive NN achieves only
  mixed results in all three experiments, whereas the stronger RNTN
  model generalizes well in every case, suggesting that it has in fact
  learned, or at least learned to simulate, our target logical
  concepts.}

@@

% TODO: Clean up and condense to column length?

>>>>

To date, experimental investigations of these concepts using
distributed representations have been largely confined to short
phrases \cite{Mitchell:Lapata:2010,Grefenstette-etal:2011,baroni2012entailment}.
However, for robust natural language understanding, it is essential to
model these phenomena in their full generality on complex linguistic
structures. Recent work on the algebraic approach of
\newcite{ClarkCoeckeSadrzadeh2011} has yielded rich frameworks 
for computing the meanings of complex fragments of natural
language compositionally from vector or tensor representations, but 
has not yet yielded effective methods for learning these representations
entirely from data in a typical machine learning setting.

@@

In our first set of experiments, we use this pre-specified logical
grammar to generate controlled data sets encoding the semantic
relationships between pairs of expressions and evaluate whether each
of two classes of neural model---plain recursive NNs and recursive neural
tensor networks (RNTNs; \citealt{socher2013acl1})---can learn those
relationships correctly. In our first experiment
(Section~\ref{sec:join}), we evaluate the ability of these models to
learn the core relational algebra of natural logic from data. Our
second experiment (Section~\ref{sec:recursion}) extends this
evaluation to relations between complex recursive structures like
$(a \plor b)$ and $\plneg (\plneg a \pland \plneg b)$, and our third
experiment (Section~\ref{sec:quantifiers}) involves relations between
quantified statements like \ii{every reptile walks} and \ii{no turtle
  moves}. We find that the plain recursive NN achieves only
  mixed results in all three experiments, whereas the stronger RNTN
  model generalizes well in every case, suggesting that it has in fact
  learned, or at least learned to simulate, our target logical
  concepts.

@@

### 2014-12-03 writing/F'14 paper/join.tex
Intro update and misc paper cleanup

In our experiments, we create 80 randomly generated sets drawing from
a domain of seven elements. This yields a data set consisting of
6400 statements about pairs of formulae. 3200 of these pairs are
chosen as a test set, and that test set is further reduced to the 2960
examples that can be provably derived from the training data.

We trained versions of both the RNN model and the RNTN model on these
data sets. In both cases, the models are implemented exactly as
described in Section~\ref{methods}, but since the items being compared
are single terms rather than full tree structures, the composition
layer is not involved, and the two models differed only in which
layer function is used for the comparison layer. We simply present
the models with two single terms, each of which corresponds to a
single vector in the (randomly initialized) vocabulary matrix $V$,
thereby ensuring that the model has no information about the terms
being compared except the relations between them.

\begin{figure}[tp]
  \centering
%  \begin{subfigure}[t]{0.45\textwidth}
%    \centering
%    \newcommand{\labelednode}[4]{\put(#1,#2){\oval(1.5,1)}\put(#1,#2){\makebox(0,0){$\begin{array}{c}#3\\\{#4\}\end{array}$}}}
%    \setlength{\unitlength}{1cm}
%    \begin{picture}(5,5.5)
%      \labelednode{2.50}{5}{}{1,2,3}
%      
%      \put(0.75,4){\line(3,1){1.5}}
%      \put(2.5,4){\line(0,1){0.5}}
%      \put(4.25,4){\line(-3,1){1.5}}
%      
%      \labelednode{0.75}{3.5}{a,b}{1,2}
%      \labelednode{2.50}{3.5}{c}{1,3}
%      \labelednode{4.25}{3.5}{d}{2,3}
%      
%      \put(0.75,2.5){\line(0,1){0.5}}
%      \put(0.75,2.5){\line(3,1){1.5}}
%      
%      \put(2.5,2.5){\line(-3,1){1.5}}
%      \put(2.5,2.5){\line(3,1){1.5}}
%      
%      \put(4.25,2.5){\line(0,1){0.5}}
%      \put(4.25,2.5){\line(-3,1){1.5}}
%      
%
%      \labelednode{0.75}{2}{e,f}{1}
%      \labelednode{2.50}{2}{}{2}
%      \labelednode{4.25}{2}{g,h}{3}
%      
%      \put(2.5,1){\line(-3,1){1.5}}
%      \put(2.5,1){\line(0,1){0.5}}
%      \put(2.5,1){\line(3,1){1.5}}
%      
%      \labelednode{2.5}{0.5}{}{}
%    \end{picture}
%    \caption{Simple boolean structure. The letters name the sets. Not all sets have names, and
%    some sets have multiple names, so that learning $\nateq$ is non-trivial.}
%  \end{subfigure}
%  \qquad
%  \begin{subfigure}[t]{0.43\textwidth}
    \centering
    \setlength{\tabcolsep}{12pt}
    \begin{tabular}[b]{c  c}
      \toprule
      Train & Test \\
      \midrule
      $b \natcov c$              & $b \nateq b$ \\
      $c \natcov d$              & $b \natcov d$ \\
      $c \natrev e$              & \strikeout{$b \natrev e$} \\
      $c \natrev g$               & \strikeout{$c \nateq f$} \\
      $e \natfor c$              & $e \natfor b$ \\
      \bottomrule
    \end{tabular}
    \caption{A train/test split of the atomic statements about the
      model.  Test statements not provable from the training data are
      crossed out.}
 % \end{subfigure}  
 %  \caption{Small example structure and data for learning relation composition.}
  \label{lattice-figure}
\end{figure} 
% Note: None of these test examples is derivable from the shown training data, 
% but we suggest that there is additional training data, so we can cross lines
% out willy-nilly without fear.

\begin{table}[tp]
  \centering \small
  \begin{tabular}{ l r@{ \ }r r@{ \ }r r@{ \ }r }
    \toprule
    ~&\multicolumn{2}{c}{$\natind$ only} & \multicolumn{2}{c}{15d RNN}  & \multicolumn{2}{c}{15d RNTN}\\
    \midrule
    Test &52.0 &(10.3) &	94.0&(87.0)& \textbf{99.6} & \textbf{(95.5)}\\
    Train &53.8 &(10.5)    & 99.8&	(99.0) & \textbf{100} & \textbf{(100)}\\

    \bottomrule
  \end{tabular}
  \caption{Performance on the semantic relation experiments. These results and all other results on artificial data are reported as mean accuracy scores over five runs followed by mean macroaveraged F1 scores in parentheses.}
  \label{joinresultstable}
\end{table}

\paragraph{Results} 
Table \ref{joinresultstable} shows the results, which are straightforward 
to interpret. The RNTN is able to accurately encode the relations 
between the terms in the geometric relations between their vectors, 
and is able to then use that information to recover relations that 
are not overtly included in the training data. The RNN, in contrast, was not able to learn
to do this well in this setting.

>>>>

In each experimental run, we create 80 randomly generated sets drawing from
a domain of seven elements. This yields a data set consisting of
6400 statements about pairs of formulae. 3200 of these pairs are
chosen as a test set, and that test set is further reduced to the 2960
examples that can be provably derived from the training data.

We trained versions of both the RNN model and the RNTN model on these
data sets. In both cases, the models are implemented exactly as
described in Section~\ref{methods}, but since the items being compared
are single terms rather than full tree structures, the composition
layer is not involved, and the two models differ only in which
layer function is used for the comparison layer. We simply present
the models with two single terms, each of which corresponds to a
single vector in the (randomly initialized) vocabulary matrix $V$,
thereby ensuring that the model has no information about the terms
being compared except the relations between them.

\begin{figure}[tp]
  \centering
%  \begin{subfigure}[t]{0.45\textwidth}
%    \centering
%    \newcommand{\labelednode}[4]{\put(#1,#2){\oval(1.5,1)}\put(#1,#2){\makebox(0,0){$\begin{array}{c}#3\\\{#4\}\end{array}$}}}
%    \setlength{\unitlength}{1cm}
%    \begin{picture}(5,5.5)
%      \labelednode{2.50}{5}{}{1,2,3}
%      
%      \put(0.75,4){\line(3,1){1.5}}
%      \put(2.5,4){\line(0,1){0.5}}
%      \put(4.25,4){\line(-3,1){1.5}}
%      
%      \labelednode{0.75}{3.5}{a,b}{1,2}
%      \labelednode{2.50}{3.5}{c}{1,3}
%      \labelednode{4.25}{3.5}{d}{2,3}
%      
%      \put(0.75,2.5){\line(0,1){0.5}}
%      \put(0.75,2.5){\line(3,1){1.5}}
%      
%      \put(2.5,2.5){\line(-3,1){1.5}}
%      \put(2.5,2.5){\line(3,1){1.5}}
%      
%      \put(4.25,2.5){\line(0,1){0.5}}
%      \put(4.25,2.5){\line(-3,1){1.5}}
%      
%
%      \labelednode{0.75}{2}{e,f}{1}
%      \labelednode{2.50}{2}{}{2}
%      \labelednode{4.25}{2}{g,h}{3}
%      
%      \put(2.5,1){\line(-3,1){1.5}}
%      \put(2.5,1){\line(0,1){0.5}}
%      \put(2.5,1){\line(3,1){1.5}}
%      
%      \labelednode{2.5}{0.5}{}{}
%    \end{picture}
%    \caption{Simple boolean structure. The letters name the sets. Not all sets have names, and
%    some sets have multiple names, so that learning $\nateq$ is non-trivial.}
%  \end{subfigure}
%  \qquad
%  \begin{subfigure}[t]{0.43\textwidth}
    \centering
    \setlength{\tabcolsep}{12pt}
    \begin{tabular}[b]{c  c}
      \toprule
      Train & Test \\
      \midrule
      $a \natfor b$              & $c \natrev a$ \\
      $b \natfor c$              & $a \natfor d$ \\
      $b \nateq d$              & \strikeout{$a \natfor e$} \\
      $d \natalt e$               & \strikeout{$e \natrev a$} \\
      $e \natalt b$              & $e \nateq e$ \\
%      a animal
%     b mammal
%      c dog
 %     d mammal'
 %     e nonmammal OR reptile
 %     b nonmammal' OR reptile'
      \bottomrule
    \end{tabular}
    \caption{A train/test split of the atomic statements about the
      model.  Test statements not provable from the training data are
      crossed out.}
 % \end{subfigure}  
 %  \caption{Small example structure and data for learning relation composition.}
  \label{lattice-figure}
\end{figure} 
% Note: None of these test examples is derivable from the shown training data, 
% but we suggest that there is additional training data, so we can cross lines
% out willy-nilly without fear.

\begin{table}[tp]
  \centering \small
  \begin{tabular}{ l r@{ \ }r r@{ \ }r r@{ \ }r }
    \toprule
    ~&\multicolumn{2}{c}{$\natind$ only} & \multicolumn{2}{c}{15d RNN}  & \multicolumn{2}{c}{15d RNTN}\\
    \midrule
    Train &53.8 &(10.5)    & 99.8&	(99.0) & \textbf{100} & \textbf{(100)}\\
    Test &52.0 &(10.3) &	94.0&(87.0)& \textbf{99.6} & \textbf{(95.5)}\\
    \bottomrule
  \end{tabular}
  \caption{Performance on the semantic relation experiments. These results and all other results on artificial data are reported as mean accuracy scores over five runs followed by mean macroaveraged F1 scores in parentheses.}
  \label{joinresultstable}
\end{table}

\paragraph{Results} 
Table \ref{joinresultstable} shows the results, which are straightforward 
to interpret. The RNTN is able to accurately encode the relations 
between the terms in the geometric relations between their vectors, 
and is able to then use that information to recover relations that 
are not overtly included in the training data. The RNN, in contrast, was not able to learn
to do this well under these conditions, though we cannot rule
out the possibility that different optimization techniques or further hyperparameter 
tuning could lead an RNN model to succeed here.

### 2014-12-03 writing/F'14 paper/methods.tex
Intro update and misc paper cleanup

We use the tree pair model architecture depicted in Figure~\ref{sample-figure}, which is novel 
but closely modeled after past work on recursive NNs. 
The two phrases being compared are processed separately using a pair 
of recursive networks that share a single set of parameters. 
The resulting vectors are fed into a separate comparison
layer (fixed at 75 dimensions) that is meant to generate a feature vector capturing the
relation between the two phrases. The output of this layer is then
given to a softmax classifier, which in turn produces a hypothesized
distribution over the seven relations represented in Table~\ref{b-table}.

For a composition layer, we evaluate models with both the plain neural
network layer function \eqref{rnn} and the more powerful RNTN layer function
\eqref{rntn} proposed in \newcite{chen2013learning}. The nonlinearity $f(x) = \tanh(x)$ is applied to the output of the layer function.
%
\begin{gather} \label{rnn}
\vec{y}_{\textit{RNN}} = f(\mathbf{M} [\vec{x}^{(l)}; \vec{x}^{(r)}] + \vec{b}\,) \\
\label{rntn}
\vec{y}_{\textit{RNTN}} = \vec{y}_{\textit{RNN}} + f(\vec{x}^{(l)T} \mathbf{T}^{[1 \ldots n]} \vec{x}^{(r)})
\end{gather} 
%
Here, $\vec{x}^{(l)}$ and $\vec{x}^{(r)}$ are the column vector
representations for the left and right children of the node, and
$\vec{y}$ is the node's output.  The RNN concatenates them, multiplies
them by an $n \times 2n$ matrix of learned weights, adds a bias $\vec{b}$, and applies the
element-wise non-linearity to the resulting vector. The RNTN adds a learned third-order tensor 
$\mathbf{T}$, of dimension $n \times n \times n$, modeling
multiplicative interactions between the child vectors.

>>>>

To apply recursive models to this inference class, we propose the tree 
pair model architecture depicted in Figure~\ref{sample-figure}, which is novel but
closely modeled after past work on recursive NNs. 
In it, the two phrases being compared are processed separately using a pair 
of recursive networks that share a single set of parameters. 
The resulting vectors are fed into a separate comparison
layer that is meant to generate a feature vector (fixed at 75 dimensions) capturing the
relation between the two phrases. The output of this layer is then
given to a softmax classifier, which in turn produces a hypothesized
distribution over the seven relations represented in Table~\ref{b-table}.

For a composition layer, we evaluate models with both the plain neural
network layer function \eqref{rnn} and the more powerful RNTN layer function
\eqref{rntn} proposed in \newcite{chen2013learning}. The nonlinearity $f(x) = \tanh(x)$ is applied to the output of the layer function.
%
\begin{gather} \label{rnn}
\vec{y}_{\textit{RNN}} = f(\mathbf{M} [\vec{x}^{(l)}; \vec{x}^{(r)}] + \vec{b}\,) \\
\label{rntn}
\vec{y}_{\textit{RNTN}} = \vec{y}_{\textit{RNN}} + f(\vec{x}^{(l)T} \mathbf{T}^{[1 \ldots n]} \vec{x}^{(r)})
\end{gather} 
%
Here, $\vec{x}^{(l)}$ and $\vec{x}^{(r)}$ are the column vector
representations for the left and right children of the node, and
$\vec{y}$ is the node's output.  The RNN concatenates them, multiplies
them by an $n \times 2n$ matrix of learned weights, adds a bias $\vec{b}$, and applies the
non-linearity element-wise to the resulting vector. The RNTN adds a learned third-order tensor 
$\mathbf{T}$, of dimension $n \times n \times n$, modeling
multiplicative interactions between the child vectors.

### 2014-12-03 writing/F'14 paper/recursion.tex
Intro update and misc paper cleanup

Complex embedding is a prominent feature of natural languages. 
The principle of compositionality explains how the meanings of 
such structures are derived systematically from their parts, thereby
 enabling people to use novel phrases and sentences.
Theoretical accounts of natural language syntax and semantics
therefore rely heavily on recursive structures. This section examines the 
question of whether our models are able to learn a 
compositional semantics over such recursive structures.
In our evaluations, we exploit the fact that our logical language
is infinite by testing on strings that are longer and more complex
than any seen in training.

@@

For our experiments, we randomly generate a large set of  unique pairs 
of formulae and compute the relation that holds for each pair.
We discard pairs in which either statement is a tautology or
contradiction (a statement that is true of either all or no possible
assignments), for which none of the seven relation labels in
Table~\ref{b-table} can hold. The resulting set of formula pairs is
then partitioned into bins based on the number of logical operators in
the longer of the two formulae. We then randomly sample 15\% of each
bin for a held-out test set.

@@

\paragraph{Results} 
We trained both the RNN and RNTN models on data with up to 4
connectives (65k pairs) and tested it on examples of up to 12
connectives (44k pairs). We initialized the model parameters randomly,
including the vector representations of the six variables. The results
are shown in Figure~\ref{prop-results}. \mynote{In tuning, we found that the
RNN model was approximately optimal with 45-dimensional vector
representations, and the RNTN model was approximately optimal with 25
dimensions. We fixed the size of the feature vector for the classifier
at 75 dimensions. We found that the RNTN model was able to perform
almost perfectly on unseen small test examples, with accuracy above
98.7\% on formulae below length four (macro-F1 was above 95.5).
Starting at size four, performance gradually falls with increasing
size.  The RNN model did not perform well, reaching only 90.5\% (81.4
macro-F1) accuracy on the smallest test examples, and declining from
there to near-baseline performance on formulae of length 12. Training
accuracy was 99.6\% for the RNTN and 87.6\% for the RNN.}

>>>>

Complex embedding is a prominent feature of natural languages. 
The principle of compositionality explains how the meanings of 
such structures are derived systematically from their parts,
 enabling people to use novel phrases and sentences.
Theoretical accounts of natural language syntax and semantics
therefore rely heavily on recursive structures. This section examines the 
question of whether our models are able to learn a 
compositional semantics over such recursive structures.
In our evaluations, we exploit the fact that our logical language
is infinite by testing on strings that are longer and more complex
than any seen in training.

@@

In our experiments, we randomly generate a large set of unique pairs 
of formulae and compute the relation that holds for each pair.
We discard pairs in which either statement is a tautology or
contradiction (a statement that is true of either all or no possible
assignments), for which none of the seven relation labels in
Table~\ref{b-table} can hold. The resulting set of formula pairs is
then partitioned into bins based on the number of logical operators in
the longer of the two formulae. We then randomly sample 20\% of each
bin for a held-out test set.

@@

\paragraph{Results} 
We trained both the RNN and RNTN models on data with up to 4
connectives (60k pairs) and tested it on a separate set of examples of up to 12
connectives (21k pairs). We initialized the model parameters randomly,
including the vector representations of the six variables, and tested the model
in five fold cross-validation, discarding all but 20\% of the examples with 
more than four connectives. The results are shown in Figure~\ref{prop-results}. 
\mynote{In tuning, we found that the
RNN model was approximately optimal with 45-dimensional vector
representations, and the RNTN model was approximately optimal with 25
dimensions. We fixed the size of the feature vector for the classifier
at 75 dimensions. We found that the RNTN model was able to perform
almost perfectly on unseen small test examples, with accuracy above
98.7\% on formulae below length four (macro-F1 was above 95.5).
Starting at size four, performance gradually falls with increasing
size.  The RNN model did not perform well, reaching only 90.5\% (81.4
macro-F1) accuracy on the smallest test examples, and declining from
there to near-baseline performance on formulae of length 12. Training
accuracy was 99.6\% for the RNTN and 87.6\% for the RNN.}
